Abstract Spatial Specification of Stencil Computations

### 2016-06-24 paper.tex
fixed spurious example and its discussion which was wrong

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteratively computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran: 
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

@@

A \term{backward} stencil
is similar but has negative neighbourhood indices, \eg{}, 
%
\begin{ExmVerbatim}
!= stencil backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{ExmVerbatim}
%
Note that the left-hand subscript on \term{b} does not affect
the stencil specification. The requirement on the left-hand side
is merely that it is an array subscript with at least one neighbourhood index,
but this may be on different induction variables to the right-hand
side, which is the stencil we are concerened with.

A \texttt{centered} stencil is a combination of \texttt{forward}
and \texttt{backward} stencils to the same depth, \eg{},
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1) :: a
b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
\end{ExmVerbatim}

@@

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by 
non-determinstically picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:  
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} = 
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5 
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors 
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise pemutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_i + \neg b_{i,j} v_i
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer $i$. 
Recall that the vectors are defined over the domain
$\mathbb{Z} \cup \{\infty\}$. 
For addition and multiplication, we take 
 $\infty$ to be an absorbing element. Thus, if either
vector in a pairwise permutation contains $\infty$ in
the $j^{th}$ entry, then $j^{th}$ entry of every pairwise
permutation is $\infty$, \eg{}, 
%%
\begin{equation*}
\vtwoh{0}{1} \bowtie \vtwoh{2}{\infty} = 
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{cc}
0 & \infty \\
0 & \infty \\ 
2 & \infty \\
2 & \infty
\end{array}
\right]
\end{equation*}

@@

\section{Correctness}

\paragraph{Proof}
\begin{itemize}
\item \textsc{coalesceF}
\begin{align*}
\llbracket & \texttt{v} : \texttt{fwd} \; \texttt{dims}=ds \; \texttt{depth}=n; \,
  \texttt{v} :  \texttt{fwd} \; \texttt{dims}=ds \;
  \texttt{depth}=n+1)\rrbracket \\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \; \cup \; \{ ix^m \mid \forall \, m, ix : \mathbb{Z}^{\mathbb{N}}, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq n \}\\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ((ix \, d \leq (n+1)) \vee (ix \,
  d \leq n)) \} \\
& =  \{ (\texttt{v}, ix)^m \mid \forall \, m, ix : \relix, d : \mathbb{N}
  \, . \, d \in \textit{ds} \Rightarrow ix \, d \leq (n+1) \} \\
& = \interp{\texttt{v} : \texttt{fwd} \; \texttt{dim}=ds \;
  \texttt{depth}=n + 1} \qquad \Box
\end{align*}
\end{itemize}

>>>>

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteratively computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:

\mnote{Line 4 does not swap  between iterations it does within. Is this 
correct? I think it needs to be on Line 5? If it is meant to be in Line 4, it is
wrongly indented.}
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

@@

A \term{backward} stencil
is similar but has negative neighbourhood indices, \eg{}, 
%
\begin{ExmVerbatim}
!= stencil backward(depth=2, dim=1) :: a
b(i) = a(i) + a(i-1) + a(i-2)
\end{ExmVerbatim}
%
A \texttt{centered} stencil is a combination of \texttt{forward}
and \texttt{backward} stencils to the same depth, \eg{},
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1) :: a
b(i) = ( a(i) + a(i-1) + a(i+1) ) / 3.0
\end{ExmVerbatim}

@@

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by 
non-determinstically picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:  
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} = 
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5 
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors 
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise pemutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer $i$. 
Recall that the vectors are defined over the domain
$\mathbb{Z} \cup \{\infty\}$. 
For addition and multiplication, we take 
 $\infty$ to be an absorbing element. Thus, if either
vector in a pairwise permutation contains $\infty$ in
the $j^{th}$ entry, then $j^{th}$ entry of every pairwise
permutation is $\infty$, \eg{}, 
%%
\begin{equation*}
\vtwoh{0}{1} \bowtie \vtwoh{2}{\infty} = 
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{cc}
0 & \infty \\
0 & \infty \\ 
2 & \infty \\
2 & \infty
\end{array}
\right]
\end{equation*}

@@

\input{proof.tex}
